---
title: "Reproducible Research: Peer Assessment 1"
author: "Tahmid Rafi"
date: "November 17, 2014"
output: 
  html_document:
    keep_md: true
    theme: journal
    highlight: pygments
---


## Setting up required libraries

```{r loadlibrary, message=FALSE}
if(!is.element("dplyr", installed.packages()[,1])){
    install.packages("dplyr")
}

library(dplyr)

if(!is.element("lattice", installed.packages()[,1])){
    install.packages("lattice")
}

library(lattice)
```


## Loading and preprocessing the data

First of all we need to unzip the data if not already done

```{r unzipdata}
data.zip <- "activity.zip"
data.file <- "activity.csv"

if(!file.exists(data.file)){
    message("Unzipping File: ", data.zip)
    unzip(zipfile = data.zip)
}

```

as it is a csv file we load the file with `read.csv()` command.

```{r loaddata}
raw.data <- read.csv(data.file)
raw.data <- tbl_df(raw.data)
raw.data
```

## What is mean total number of steps taken per day?

Let's take a look at the mean total number of steps taken per day.
```{r summarizedata}
steps.per_day <- raw.data %>% 
  select(steps, date) %>% 
  group_by(date) %>% 
  summarize(total.steps = sum(steps)) %>% 
  arrange(as.Date(date))

head(steps.per_day, n=10)
```

Now let's look at the histogram of the data

```{r drawhistogram}
hist(steps.per_day$total.steps, 
     xlab = "Total no of steps", 
     main = "Histogram of total number of steps", 
     col = "grey", breaks = seq(0,25000,1000))
```

The mean and the median of total step taken per day are as follows.

```{r calcutaleMeanMedian}
mean.step <- mean(steps.per_day$total.steps, na.rm = TRUE)
print(paste("The mean total number of steps taken per day is", format(mean.step, nsmall = 2, digits = 2)))

median.step <- median(steps.per_day$total.steps, na.rm = TRUE)
print(paste("The median total number of steps taken per day is", format(median.step, nsmall = 2, digits = 2)))
```


## What is the average daily activity pattern?

Now, Let us analyze the average daily activity pattern. 

```{r stepPerInterval}
steps.per_interval <- raw.data %>% 
  select(steps, interval) %>% 
  group_by(interval) %>% 
  summarize(mean.steps = mean(steps, na.rm = TRUE))
```

Let's draw a plot

```{r drawSPI}
plot(steps.per_interval, type= "l", 
     xlab = "5-minute interval", ylab = "Avg. steps", 
     main = "Daily Activity Pattern", col = "red")
```

Let's find out which time of the day or which interval has the most no of steps, averaged across all days?

```{r mostSteppedInterval}
mostStepped <- steps.per_interval %>% arrange(desc(mean.steps))
print(mostStepped)
```

Which shows that the most stepped interval is **`r mostStepped$interval[1]`**


## Imputing missing values

#### Number of missing values

```{r naCount}
sum(is.na(raw.data))
```

Analyzing two previous graph's we can see that there are days where all the values are `NA`'s. So, if we use daily means to impute the NA values, then we will have some `NaN`'s. But in the Average per Interval their are no NA's. So this is a good option to fill out the NA's by average steps per Interval.   Let's fill out the `NA`'s with the mean value for the interval averages across all days.

```{r imputeNA}
imputed_data <- raw.data
for (i in 1:nrow(imputed_data))
{
  if( is.na(imputed_data[i,][['steps']]) ) {
    interval = imputed_data[i,][['interval']]
    minute <- (interval %/% 100) * 60 + (interval %% 100)
    index = minute / 5 + 1
    imputed_data[i,][['steps']] = steps.per_interval[index,][['mean.steps']]
    
  }
}

summary(imputed_data)
```

Let's Make another Histogram from the imputed data

```{r summarizeNewData}
New_steps.per_day <- imputed_data %>% 
  select(steps, date) %>% 
  group_by(date) %>% 
  summarize(total.steps = sum(steps)) %>% 
  arrange(as.Date(date))

head(New_steps.per_day, n=10)
```

```{r drawNewHistogram}
hist(New_steps.per_day$total.steps, 
     xlab = "Total no of steps", 
     main = "Histogram of total number of steps", 
     col = "grey", breaks = seq(0,25000,1000))
```

```{r calcutaleNewMeanMedian}
mean.new.step <- mean(New_steps.per_day$total.steps)
print(paste("The mean total number of steps taken per day is", format(mean.new.step, nsmall = 2, digits = 2)))

median.new.step <- median(New_steps.per_day$total.steps)
print(paste("The median total number of steps taken per day is", format(median.new.step, nsmall = 2, digits = 2)))
```


## Are there differences in activity patterns between weekdays and weekends?

Modify the data to have a new column based on "weekday" and "weekend"

```{r weekData}
week_data <- imputed_data %>% 
  select(steps, date, interval) %>% 
  mutate(date = as.Date(date)) %>% 
  mutate(day = ifelse(weekdays(date) %in% c("Friday", "Saturday"), "weekend", "weekday")) %>%
  mutate( day = factor(day))
```

Calculate Number of Steps per day

```{r Newsteps}
weekdayStep <- week_data %>% 
  select(steps, interval, day) %>% 
  filter(day == "weekday") %>%
  group_by(interval) %>% 
  summarize(total.steps = mean(steps)) %>%
  mutate(day = "weekday")

weekendStep <- week_data %>% 
  select(steps, interval, day) %>% 
  filter(day == "weekend") %>%
  group_by(interval) %>% 
  summarize(total.steps = mean(steps)) %>%
  mutate(day = "weekend")

total_step <- rbind(weekdayStep,weekendStep) %>%
  mutate(day = factor(day))

xyplot(total.steps ~ interval | day, 
       data = total_step, type='l',
       layout = c(1,2))
```



