---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions}
library(knitr)
opts_chunk$set(digits = 2)
```

## Setting up required libraries

```{r loadlibrary}
if(!is.element("dplyr", installed.packages()[,1])){
    install.packages("dplyr")
}

library(dplyr)
```

Load and set options on knitr

```{r}
if(!is.element("knitr", installed.packages()[,1])){
    install.packages("knitr")
}

library(knitr)

```


## Loading and preprocessing the data

First of all we need to unzip the data if not already done

```{r unzipdata}
data.zip <- "activity.zip"
data.file <- "activity.csv"

if(!file.exists(data.file)){
    message("Unzipping File: ", data.zip)
    unzip(zipfile = data.zip)
}

```

as it is a csv file we load the file with `read.csv()` command.

```{r loaddata}
raw.data <- read.csv(data.file)
summary(raw.data)
```

## What is mean total number of steps taken per day?

Let's take a look at the mean total number of steps taken per day.
```{r summarizedata}
steps.per_day <- raw.data %>% 
  select(steps, date) %>% 
  group_by(date) %>% 
  summarize(total.steps = sum(steps)) %>% 
  arrange(as.Date(date))

head(steps.per_day, n=10)
```

Now let's look at the histogram of the data

```{r drawhistogram}
hist(steps.per_day$total.steps, 
     xlab = "Total no of steps", 
     main = "Histogram of total number of steps", 
     col = "grey", breaks = seq(0,25000,2500))
```

```{r calcutaleMeanMedian}
mean.step <- mean(steps.per_day$total.steps, na.rm = TRUE)
print(paste("The mean total number of steps taken per day is", format(mean.step, nsmall = 2, digits = 2)))

median.step <- mean(steps.per_day$total.steps, na.rm = TRUE)
print(paste("The median total number of steps taken per day is", format(median.step, nsmall = 2, digits = 2)))
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
